---
title: "RNAseq Analysis"
author: "Eirik Høye"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(DESeq2)
library(readxl)
library(sva)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(tximportData)
library(tximport)
library(GenomicFeatures)
library(RColorBrewer)
library(rtracklayer)

```

```{r}
proj_path <- '~/Dropbox/deconv_metastasis/'
```

Load Metadata

```{r}
metadata <- read_tsv("~/Dropbox/deconv_metastasis/data/metadata/metadata_v2.txt")
head(metadata)
```

Import GTF File and Extract Gene Information

```{r}
dir <- system.file("extdata", package = "tximportData")
gtf <- import(file.path(dir, 'gencode.v46.chr_patch_hapl_scaff.annotation.gff3.gz'))
protein_coding_genes <- gtf[gtf$gene_type == "protein_coding"]

txdb <- makeTxDbFromGRanges(protein_coding_genes)
gene_info <- data.frame(
  gene_id = mcols(protein_coding_genes)$gene_id,
  gene_name = mcols(protein_coding_genes)$gene_name
)
gene_info <- unique(gene_info)
head(gene_info)
```

Create tx2gene Mapping

```{r}
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- select(txdb, k, "GENEID", "TXNAME")
head(tx2gene)
```

Load and Import Count Data
GCF0506

```{r}
dir_GCF0506 <- paste0(proj_path, "data/counts/data_for_Eirik/GCF0506")
samples_GCF0506 <- read.table(file.path(dir_GCF0506, "sample_info_updated.txt"), header=TRUE)
files_GCF0506 <- file.path(dir_GCF0506, "tx_import", samples_GCF0506$name, "quant.sf")
names(files_GCF0506) <- samples_GCF0506$name
all(file.exists(files_GCF0506))
txi_GCF0506 <- tximport(files_GCF0506, type = "salmon", tx2gene = tx2gene)
GCF0506_counts <- as.data.frame(txi_GCF0506$counts)
replicate_names <- colnames(GCF0506_counts)
```

GCF0620

```{r}
dir_GCF0620 <- paste0(proj_path, "data/counts/data_for_Eirik/GCF0620")
samples_GCF0620 <- read.table(file.path(dir_GCF0620, "sample_info.txt"), header=TRUE)
files_GCF0620 <- file.path(dir_GCF0620, "tx_import", samples_GCF0620$name, "quant.sf")
names(files_GCF0620) <- samples_GCF0620$name
all(file.exists(files_GCF0620))
txi_GCF0620 <- tximport(files_GCF0620, type = "salmon", tx2gene = tx2gene)
GCF0620_counts <- as.data.frame(txi_GCF0620$counts)
replicate_names <- colnames(GCF0620_counts)

```

Merge Count Matrices

```{r}
counts <- merge(GCF0620_counts, GCF0506_counts, by=0)
rownames(counts) <- counts$Row.names
counts$Row.names <- NULL
```

Prepare Metadata for DESeq2

```{r}
coldata <- as.data.frame(metadata)
coldata <- coldata[order(rownames(coldata)), ]
coldata$tissue_type <- as.factor(coldata$tissue_type)
coldata$batch <- as.factor(coldata$batch)
coldata$TC_rank <- as.factor(coldata$TC_rank)
coldata_smp <- rbind(samples_GCF0620, samples_GCF0506)
coldata_smp$smp <- str_replace(coldata_smp$smp, '-', '_')
coldata_smp$smp <- str_replace(coldata_smp$smp, '-', '_')
coldata <- left_join(coldata_smp, coldata, by=c('smp' = 'sample_name'))
rownames(coldata) <- coldata$name
coldata$name <- NULL

# remove the samples that were not derived from colorectal cancer
only_include <- rownames(coldata %>% filter(!smp %in% c("PCB_714", "PC_301")))
counts <- counts[only_include]
coldata <- coldata %>% filter(!smp %in% c("PCB_714", "PC_301"))
```

```{r}
print(dim(coldata))
```

```{r}
print(dim(counts))
```

DESeq2 Analysis

```{r}
dds <- DESeqDataSetFromMatrix(countData = round(counts),
                              colData = coldata,
                              design = ~ tissue_type)
dds <- DESeq(dds)
summary(dds)
```

Visualize PCA

```{r}
vst <- vst(dds, blind=FALSE)
plotPCA(vst, intgroup=c('tissue_type', 'batch'))
```

Collapse Replicates and Perform PCA

```{R}
dds$run <- colnames(dds)
ddsColl <- collapseReplicates(dds, dds$smp, dds$run)
vstColl <- vst(ddsColl, blind=FALSE)
plotPCA(vstColl, intgroup=c('tissue_type', 'batch'))
```

Compae PM datasets from different sequencing runs

```{r}
ddsPM <- DESeqDataSetFromMatrix(countData = round(counts), colData = coldata, design = ~ batch)
ddsPM <- ddsPM[, ddsPM$tissue_type == 'PM']
ddsPM$run <- colnames(ddsPM)
ddsPMColl <- collapseReplicates(ddsPM, ddsPM$smp, ddsPM$run)
ddsPMColl <- DESeq(ddsPMColl)
vstPMColl <- vst(ddsPMColl, blind=FALSE)

```

PCA and Differential Gene Expression Analysis

```{r}
vstPMColl_PCA <- PCA(t(assay(vstPMColl)), graph=FALSE)

axes <- c(1,2)
vstPMColl$batch <- factor(vstPMColl$batch, levels = c("GCF0620", "GCF0506"))

p1 <- fviz_pca_ind(vstPMColl_PCA,
             axes = axes,
             palette = 'aaas',
             geom.ind = 'point',
             pointshape = 21,
             pointsize = 2.5,
             fill.ind = vstPMColl$batch,
             col.ind = 'black'
            ) +
  ggtitle("") +
  labs(fill="Sequencing run") +
  xlab("Dimension 1") +
  ylab("Dimension 2") +
  theme(legend.position = "bottom")

alpha_level <- 0.1

res <- results(ddsPMColl, name="batch_GCF0620_vs_GCF0506", alpha=alpha_level)
resOrdered <- res[ order(res$padj) , ]
resSig <- subset(resOrdered, padj < alpha_level)

p2 <- resOrdered %>% as.data.frame() %>%
  mutate(Gene = rownames(.)) %>% as_tibble() %>%
  mutate(Significant = case_when( padj < alpha_level ~ 'yes',
                                  TRUE ~ 'no'
                                )) %>%
  mutate(sig_lab = case_when( padj < alpha_level & abs(log2FoldChange) > 3 ~ Gene,
                              TRUE ~ '')) %>%
  ggplot(aes(x=log2FoldChange, y=-log10(padj), fill = Significant)) +
  geom_point(shape=21) +
  theme_minimal() +
  geom_hline(yintercept = -log10(alpha_level), linetype = 'dashed',
             color = 'red') +
  ggtitle("") +
  theme(plot.title = element_text(face = "bold"), legend.position = "none") +
  scale_fill_manual(values=c("black", "red"))
p <- ggarrange(p1, p2)
p
```


```{r}
summary(res)

```

Top 15 Differentially Expressed Genes

```{r}
set <- resOrdered[1:15, ] 
set['Gene_name'] = rownames(set)
set <- set %>% as.tibble()
set <- transform(set, lower_bound = log2FoldChange - 1.96 * lfcSE,
                       upper_bound = log2FoldChange + 1.96 * lfcSE) %>% 
  merge(gene_info, by.x='Gene_name', by.y='gene_id') %>% arrange(padj) %>%
  mutate(padj = as.factor(padj))
```

Bar Plot of Top Genes

```{r}

colors <- c(brewer.pal(12, "Paired"), brewer.pal(3, "Set2"))

set %>%
  ggplot(aes(x=log2FoldChange, y=Gene_name, fill=gene_name)) +
  geom_bar(stat = 'identity') +
  geom_errorbarh(aes(xmin = lower_bound, xmax = upper_bound), height = 0) +
  scale_fill_manual(values = colors) +
  theme_classic()
```








